Fast and scalable Gaussian process modeling of stellar variability \vspace{-3\baselineskip}

### 2016-12-19 paper/ms.tex

%The general ACF is given by
%\begin{equation}
%K(\tau) = a_r e^{-b_r \vert\tau\vert} \cos{(b_r \vert\tau\vert)} + a_i e^{-b_r \vert\tau\vert} \sin{(b_r \vert\tau\vert)}.
%\end{equation}
%which has a power spectrum given by
%\begin{equation}
%P(\omega) = P_0 \frac{(a_ib_i+a_rb_r)(b_i^2+b_r^2)+(a_rb_r-a_ib_i)\omega^2}{(\omega^2-(b_i^2+b_r^2))^2 + 4b_r^2\omega^2}.
%\end{equation}
%The power spectrum should be non-negative since it is the absolute square of the Fourier transform of the noise.
%Note that the denominator is always non-negative, so the power spectrum is non-negative if $a_ib_i+a_rb_r \ge 0$ and
%$a_rb_r-a_ib_i \ge 0$.

@@

In the limit of an infinite time series and white random forcing, the noise-free power 
spectrum of this equation is given by:
\begin{equation}
P(\omega) = P_0 \frac{\omega_0^4}{(\omega^2-\omega_0^2)^2 + \omega_0^2\omega^2/Q^2},
\end{equation}
where $P_0$ is a normalization constant.

% Check to see that signs match
This power spectrum matches that of our basis kernel if
\begin{eqnarray}
\omega_0^2 &=& c_j^2+d_j^2,\\
a_jc_j-b_jd_j &=& 0,\\
\omega_0^2&=& b_jd_j+a_jc_j ,\\
\omega_0^2/Q^2&=& 4c_j^2.
\end{eqnarray}
The power spectrum can be normalized by $P_0$ in amplitude and $\omega_0$ in
frequency to yield a spectral shape that only depends upon $Q$.  This is plotted
in Figure \ref{sho_psd} for various values of $Q$.  
Note that the oscillation frequency that appears in the kernel, $d_j$, differs
from that of the simple-harmonic oscillator, $\omega_0$, due to the damping
frequency, $c_j$.  This power spectrum has
a peak frequency given by $\omega_{max} = \omega_0\sqrt{1-1/(2Q^2)} = d_j$, which is
greater than zero for $Q > 1/\sqrt{2}$.

\begin{figure}[!htbp]
\begin{center}
\includegraphics[width=\textwidth]{./approximate_kernels/sho.png}
\caption{Normalized power spectra of the stochastically-driven, damped simple
harmonic oscillator.}
\label{sho_psd}
\end{center}
\end{figure}

This power spectrum has several limits of interest:
\begin{itemize}
\item $Q=1/2$:  This results in a power spectrum $\propto 1/(x^2+1)^2$, where $x=\omega/\omega_0$.
This matches the power spectrum of the Mat\'ern 3/2 kernel, and can be represented by the following
autocovariance function:
\begin{equation}
K(\tau) = a_j e^{-c_j \vert\tau\vert} \left(\cos{(d_j \vert\tau\vert)} + \frac{c_j}{d_j}\sin{(d_j \vert\tau\vert)}\right),
% Check to see that signs match in this equation.
\end{equation}
in the limit that $b_i \rightarrow \infty$, for which the second term has a coefficients that diverges,
but with large values of $b_i$, this may approximate the Mat\'ern kernel well.  Note that this
power spectrum peaks at $\omega=0$.
\item $Q=1/\sqrt{2}$:  This results in a power spectrum $\propto 1/(x^4+1)$, which is a functional
form used to fit the Solar granulation power spectrum by \cite{2009A&A...495..979M}.  This kernel has
dependence
\begin{equation}
K(\tau) = a_r e^{-b_r\vert\tau\vert} \left(\cos{(b_i\vert\tau\vert)}+\sin{(b_i\vert\tau\vert)}\right).
\end{equation}
Despite the sinusoidal dependence of the kernel, the strong damping causes the power spectrum to
peak at zero frequency; such a low quality oscillator doesn't display even quasi-periodic behavior.
\item $Q \gg 1$: This represents a high-quality oscillator which approaches a narrow peak.  The
power spectrum approaches
\begin{equation}
P(\omega) \approx P_0 \frac{Q^2}{1-1/(4Q^2)+4x_{max}^2\left(x-x_{max}\right)^2Q^2},
\end{equation}
where $x_{max}=\sqrt{1-1/(2Q^2)}$.  This shape approaches a Lorentzian for $x \approx x_{max}$ with
a width of $1/(2Qx_{max})$ and a peak height $\propto Q^2$.
\end{itemize}

The stochastically-driven, damped simple-harmonic oscillator can then represent a wide range of intrinsic 
stellar variability, with low $Q \approx 1$ representing granulation noise, and high $Q >> 1$ representing 
asteroseismic variability.

@@

Relation of positive power spectrum to positive definite covariance matrix.
Power spectrum of a sum of kernels.
Sturm's theorem applied to numerator of power spectrum.

@@

The kernel is modeled as a sum of damped harmonics,
\begin{equation}
K(\tau) = a_0 \delta(\tau) + \sum_{j=1}^{N_{kernel}} a_j e^{-\beta_R \tau} \cos{(\beta_I \tau)},
\end{equation}
where $\tau > 0$ is a time-offset between data points, $\beta_R$ represents the decay
in correlation with time and $\beta_I$ represents oscillations of period $P = 2\pi/\beta_I$, and
$a_0 = \sigma^2$ is a white-noise component with $\delta(0) = 1$ and zero otherwise.  Formally
the $\delta(\tau)$ function is a Kronecker delta function for which $\tau$ is the midpoint in
time measured over a finite duration exposure time, and $\sigma$ is the standard deviation
of the measured times if the correlated components are zero.

@@

A stationary, smooth kernel may generally be expressed as a Taylor series for the $\tau \ne 0$ component,
plus the white noise component:
\begin{equation}
K(\tau) = a_0 \delta(\tau) +  \sum_{j=0}^\infty c_j \vert\tau\vert^j.
\end{equation}
Note that this encompasses all smooth, time-stationary kernels, and generally $c_1 \le 0$ as
correlated noise initially declines with separation in time.  We wish to approximate this
as the sum of damped cosine kernels, and since this general kernel is symmetric in $\tau$, we
can carry out the approximation in terms of sums of damped cosine kernel components
\begin{equation}
K_j(\tau,a_j,\beta_{R,j},\beta_{I,j}) = a_j \exp{(\beta_{R,j} \tau)}\cos{(\beta_{I,j} \tau)},
\end{equation}
which has a Taylor series expansion:
\begin{eqnarray}
\exp{(\beta_R \tau)}\cos{(\beta_I \tau)} &=& 1 - \beta_R \tau + \frac{1}{2}\left(\beta_R^2 - \beta_I^2\right)\tau^2 \\c &-&
\frac{1}{6} \left(\beta_R^3-3\beta_R\beta_I^2\right)\tau^3 + \frac{1}{24} \left(\beta_R^4+\beta_I^4-6\beta_R^2\beta_I^2\right) \tau^4+ O(\tau^5),
\end{eqnarray}
which may be continued to any order in $\tau$ as necessary.
If an approximation to $K(\tau)$ is desired in a Taylor series to some high order $N$, this
requires combining $N/2$ components of $K_j(\tau,a_j,\beta_{R,j},\beta_{I,j})$ with $\beta_{R,j}$ and $\beta_{I,j}$
chosen to match the coefficients $c_j$.  Note that for $N=1$ we only require a damped exponential
kernel with $\beta_I = 0$.

@@

\item {\bf Exponential kernel.} This kernel, $K(\tau) = e^{-\vert \tau/\tau_0\vert}$, is exactly

@@

\item {\bf Cosine kernel.} The cosine kernel, $K(t) = \cos{\left(\frac{2\pi}{P} \vert \tau \vert\right)}$,

@@

Precition/interpolation involves constructing a matrix of covariances
between the (noisy) training set and the times at which we wish to
predict or interpolate data points.   If our training times are given by
${\bf t}$, while the predicted times are given by ${\bf t}_*$, then
mean values at the predicted times are given by (RW 2.23):
\begin{eqnarray}
\bar f_{*,k} &=& \sum_j K(t_{*,k},t_j) b_j,\\
&=& \sum_j \sum_p \alpha_p e^{-\beta_p \vert t_{*,k}-t_j\vert} b_j,\\
&=& \sum_p \alpha_p \sum_j e^{-\beta_p \vert t_{*,k}-t_j\vert} b_j,\\
y_k &=& \sum_j K(t_j,t_k) b_j,
\end{eqnarray}
where the latter equation may be solved for ${\bf b}$ using the extended
matrix formalism.  Let the length of ${\bf t}_*$ be $M$; then, the
matrix $K({\bf t}_*,{\bf t})$ has a size $M \times N$, and so a total
of $MN$ multiplications is required to obtain the predicted values.
If $M$ is large, this can become quickly prohibitive.  It turns out
that the structure of the sum of exponential kernels may be exploited
to obtain the predicted mean in of order $p(M+N)+p^2N$ multiplications,
i.e.\ a linear scaling with the total number of data points, which can
be much fewer if $M$ is large.

@@

The variance may be handled in a similar manner, but with $p^2$ coefficients
to recursively update. For example, if $t_{*,k} > t_j \forall j,k$ then:
\begin{eqnarray}
cov(f_{*,k}) &=&  K(t_{*,k},t_{*,k}) - \sum_p \sum_q \alpha_p \alpha_q e^{-\beta_p(t_{*,k}-{\bf t}^T)} K({\bf t},{\bf t})^{-1} e^{-\beta_q (t_{*,k}-{\bf t})},\\
&=& K(0) - \sum_p \sum_q \alpha_p \alpha_q \zeta_{p,q,k},\\
\zeta_{p,q,k} &=&   e^{-\beta_p(t_{*,k}-{\bf t}^T)} K({\bf t},{\bf t})^{-1} e^{-\beta_q (t_{*,k}-{\bf t})},\\
&=&  e^{-(\beta_p+\beta_q)(t_{*,k}-t_{*,k-1})} \zeta_{p,q,k-1}.
\end{eqnarray}
The recursive updates to the $\zeta_{p,q,k}$ parameters can be made at each step.
\end{itemize}

@@

Here we adapt the generalized Rybicki-Press (GRP) algorithm to handle periodic kernels.  We
follow closely the notation and equations given in \citet{Ambikasaran:2015}.  We start
with the observation that a Lorentzian correlation function may be decomposed into the sum of two
exponential functions:
\begin{eqnarray}
K(\tau) &=& \alpha e^{-\beta_R \tau} \cos{(\omega \tau)}\\
 &=& \frac{1}{2}\alpha \left[e^{-\beta \tau}+e^{-\beta^* \tau}\right],
\end{eqnarray}
where $\beta = \beta_R + i\beta_I$ is the
complex kernel parameter (i.e.\ $\beta_R = Re(\beta)$, $\beta_I = Im(\beta)$), $\omega = -\beta_I = 2\pi/P$
is the frequency of oscillation of the kernel with period $P$, and $x^*$ is the complex conjugate of $x$.

@@

\begin{figure}[!htbp]
\begin{center}
\includegraphics[scale=0.175]{./twoterm.pdf}
\hspace{0.2in}
\includegraphics[scale=1.0]{./colorcode.pdf}
\end{center}
\caption{Pictorial description of the extended sparse matrix where $N=10$, $p_0=0$, and $p=2$, following
the example shown in \citet{Ambikasaran:2015}.}
\label{matrix_structure}
\end{figure}

>>>>

%The general ACF is given by
%\begin{equation}
%k(\tau) = a_r e^{-b_r \tau} \cos{(b_r \tau)} + a_i e^{-b_r \tau} \sin{(b_r \tau)}.
%\end{equation}
%which has a power spectrum given by
%\begin{equation}
%P(\omega) = P_0 \frac{(a_ib_i+a_rb_r)(b_i^2+b_r^2)+(a_rb_r-a_ib_i)\omega^2}{(\omega^2-(b_i^2+b_r^2))^2 + 4b_r^2\omega^2}.
%\end{equation}
%The power spectrum should be non-negative since it is the absolute square of the Fourier transform of the noise.
%Note that the denominator is always non-negative, so the power spectrum is non-negative if $a_ib_i+a_rb_r \ge 0$ and
%$a_rb_r-a_ib_i \ge 0$.

@@

In the limit of an infinite time series and white random forcing, the noise-free power 
spectrum of this equation is given by:
\begin{equation}
P(\omega) = P_0 \sqrt{\frac{2}{\pi}} \frac{\omega_0^4}{(\omega^2-\omega_0^2)^2 + \omega_0^2\omega^2/Q^2},
\end{equation}
where $P_0$ is a normalization constant.

% Check to see that signs match
This power spectrum matches that of our basis kernel if
\begin{eqnarray}
a_j &=& P_0  \omega_0 Q\\
b_j &=& a_j \left(4Q^2-1\right)^{-1/2}\\
c_j &=& \frac{\omega_0}{2Q}\\
d_j &=& \omega_0 \left(1-\frac{1}{4Q^2}\right)^{1/2},
%\omega_0^2 &=& c_j^2+d_j^2,\\
%a_jc_j-b_jd_j &=& 0,\\
%\omega_0^2&=& b_jd_j+a_jc_j ,\\
%\omega_0^2/Q^2&=& 4c_j^2.
\end{eqnarray}
for $Q \ge \frac{1}{2}$ and $j=p=1$.  This gives a kernel with dependence
\begin{eqnarray}\eqlabel{sho_kernel}
%k(\tau) &=& P_0  \omega_0 Q e^{-\frac{\omega_0 \tau}{2Q}} \left[\cos{\lambda}+\frac{1}{\left(4Q^2-1\right)^{1/2}}
k(\tau) &=&  P_0  \omega_0 Q e^{-\frac{\omega_0 \tau}{2Q}} \cos{\left(\lambda-\phi\right)},\\
\lambda &=& \left(1-\frac{1}{4Q^2}\right)^{1/2}\omega_0\tau,\\
\tan{\phi} &=& \left(4Q^2-1\right)^{-1/2}.
\end{eqnarray}
The power spectrum can be normalized by $P_0\omega_0 Q$ in amplitude and $\omega_0$ in
frequency to yield a spectral shape that only depends upon $Q$.  This is plotted
in Figure \ref{sho_psd} for various values of $Q$.  
Note that the oscillation frequency that appears in the kernel, $d_j$, differs
from that of the simple-harmonic oscillator, $\omega_0$, due to the damping
frequency, $c_j$.  This power spectrum has
a peak frequency given by $\omega_{max} = \omega_0\sqrt{1-1/(2Q^2)} = d_j$, which is
greater than zero for $Q > 1/\sqrt{2}$.

\begin{figure}[!htbp]
\begin{center}
\includegraphics[width=0.75\textwidth]{./approximate_kernels/sho.png}
\caption{Normalized power spectra of the stochastically-driven, damped simple
harmonic oscillator. The dotted black line shows the peak frequency, $d_j = \omega_0 (1-1/(2Q)^2)^{1/2}$.}
\label{sho_psd}
\end{center}
\end{figure}

This power spectrum has several limits of interest:
\begin{itemize}
\item $Q=1/2$:  This corresponds to a critically damped harmonic oscillator, which results in a power 
spectrum $\propto 1/(x^2+1)^2$, where $x=\omega/\omega_0$.
This matches the power spectrum of the Mat\'ern 3/2 kernel, and can be represented by the following
autocovariance function:
\begin{equation}
k(\tau) = a_j e^{-c_j \tau} \left[\cos{(d_j \tau)} + \frac{c_j}{d_j}\sin{(d_j \tau)}\right],
% Check to see that signs match in this equation.
\end{equation}
in the limit that $d_j \rightarrow \infty$, corresponding to $\phi=\pi/2$, for which the second 
term has a coefficient that diverges, but with large values of $d_j$, this may approximate the 
Mat\'ern kernel well.  Note that this power spectrum peaks at $\omega=0$.
\item $Q=1/\sqrt{2}$:  This results in a power spectrum $\propto 1/(x^4+1)$, which is a functional
form used to fit the Solar granulation power spectrum by \cite{2009A&A...495..979M}.  This kernel has
dependence
\begin{equation}
k(\tau) = a_j e^{-c_j\tau} \left[\cos{(d_j\tau)}+\sin{(d_j\tau)}\right],
\end{equation}
corresponding to $\phi = \pi/4$.
Despite the sinusoidal dependence of the kernel, the strong damping causes the power spectrum to
peak at zero frequency; such a low quality oscillator doesn't even display quasi-periodic behavior.
\item $Q \gg 1$: This represents a high-quality oscillator which approaches a narrow peak. 
The kernel approaches
\begin{equation}
k(\tau) = P_0 \omega_0 Q e^{-\omega_0\tau/(2Q)} \cos{\left(\omega_0\tau\right)},
\end{equation}
for $Q \gg 1$.  The power spectrum approaches
\begin{equation}
P(\omega) \approx P_0 \frac{Q^2}{1-1/(4Q^2)+4x_{max}^2\left(x-x_{max}\right)^2Q^2},
\end{equation}
where $x_{max}=\sqrt{1-1/(2Q^2)}$.  This shape approaches a Lorentzian for $x \approx x_{max}$ with
a width of $1/(Qx_{max})$ and a peak height $\propto Q^2$.
\end{itemize}

The stochastically-driven, damped simple-harmonic oscillator can then describe a wide range of intrinsic 
stellar variability, with low $Q \approx 1$ representing granulation noise, and high $Q \gg 1$ representing 
asteroseismic variability.  Taking the sum over oscillators with various $Q, P_0,$ and $\omega_0$ can
give an accurate accounting of the power spectrum of stellar variability with $Q \ge \frac{1}{2}$ and
$P_0 >0$.  As this kernel is exactly described by the exponential kernel, this functional form
is tractable to solving in $\mathcal{O}(N)$ operations, as described next.

@@

The power spectrum is computed from taking the square of the Fourier transform of the data
time series; hence, the power-spectrum must be non-negative.  In constructing a kernel,
if any of the parameters $a_j$ or $b_j$ are negative, then it is possible for the power
spectrum to go negative, which violates the definition of the power spectrum.  Consequently,
if any of these coefficients is negative, it is necessary to check that the power spectrum
is still non-negative.  Note that a non-negative power-spectrum does not require the auto-correlation
function to be positive for all $\tau$.  The positive power spectrum is related to the
positive eigenvalues of the covariance matrix which are necessary for a positive-definite matrix
in limiting cases \citep{messerschmitt2006autocorrelation}.  We find empirically that requiring an everywhere
positive power spectrum results in positive eigenvalues for the covariance matrix, and
so we describe here how to ensure a positive power spectrum using Sturm's theorem.
%Relation of positive power spectrum to positive definite covariance matrix.
%Power spectrum of a sum of kernels.

In the case of $p$ complex Lorentzians we can check for negative values of the PSD by solving for the
roots of the power spectrum, abbreviating with $z = \omega^2$:
\begin{equation}
P(\omega)=  \sum_{j=1}^{p} \frac{q_j z + r_j}{z^2+s_jz + t_j} = 0
\end{equation}
where
\begin{eqnarray}
q_j &=& a_jc_j-b_jd_j\\
r_j &=& (d_j^2+c_j^2)(b_jd_j+a_jc_j)\\
s_j &=& 2(c_j^2-d_j^2)\\
t_j &=& (c_j^2+d_j^2)^2.
\end{eqnarray}
The denominators of each term are positive, so we can multiply through by $\Pi_j \left(z^2+s_jz + t_j\right)$ yielding:
\begin{equation}
P_0(z) = \sum_{j=1}^{p} (q_j z + r_j)\Pi_{k \ne j}\left(z^2+s_kz + t_k\right) = 0,
\end{equation}
which is a polynomial with order $p_{ord}=2(p-1)+1$.  With $p=2$, this yields a cubic equation
which may be solved exactly for the roots.

A procedure based upon Sturm's theorem \citep{Dorrie1965} allows one to determine whether there are any 
real roots within the range $(0,\infty]$. We first construct $P_0(z)$ and it's derivative 
$P_1(z) = P^\prime(z)$, and then loop from $k=2$ to $k=p_{ord}$, computing 
$P_k(z) = -{\rm rem}(P_{k-2},P_{k-1})$.  The function ${\rm rem}(p,q)$ is the remainder polynomial after dividing $p(z)$ by $q(z)$.

We evaluate the $z^0$ coefficients of each of the polynomial in the series by evaluating $f_0 = \{P_0(0),...,P_{p_{ord}}(0)\}$ to
give us the signs of these polynomials evaluated at $z=0$.
Likewise, we evaluate the coefficients of the largest order term in each polynomial which gives the sign of the polynomial 
as $z \rightarrow \infty$.  This gives $f_\infty = \{C(P_0,p_{ord}),C(P_1,p_{ord}-1),..., C(P_{p_{ord}},1)\}$ where 
$C(p(z),m)$ returns the coefficient of $z^m$ in polynomial $p(z)$.

With the series of coefficients $f_0$ and $f_\infty$, we then determine how many times the sign changes in each 
of these, where $\sigma(0)$ is the number of sign changes at $z=0$, and $\sigma(\infty)$ is the number of sign 
changes at $z \rightarrow \infty$.  The total number of real roots in the range $(0,\infty]$ is given 
by $N_{+}=\sigma(0)-\sigma(\infty)$.

We have checked that this procedure works for a wide range of parameters, and we find that it robustly 
matches the number of positive real roots which we evaluated numerically.

The advantage of this procedure is that it does not require computing the roots, but only carrying out algebraic 
manipulation of polynomials to determine the number of positive real roots.  If a non-zero real root is found, then
likelihood may be set to zero.
%Sturm's theorem applied to numerator of power spectrum.

@@

The kernel is modeled as a sum of damped harmonics,
\begin{equation}
k(\tau) = a_0 \delta(\tau) + \sum_{j=1}^{N_{kernel}} a_j e^{-\beta_R \tau} \cos{(\beta_I \tau)},
\end{equation}
where $\tau > 0$ is a time-offset between data points, $\beta_R$ represents the decay
in correlation with time and $\beta_I$ represents oscillations of period $P = 2\pi/\beta_I$, and
$a_0 = \sigma^2$ is a white-noise component with $\delta(0) = 1$ and zero otherwise.  Formally
the $\delta(\tau)$ function is a Kronecker delta function for which $\tau$ is the midpoint in
time measured over a finite duration exposure time, and $\sigma$ is the standard deviation
of the measured times if the correlated components are zero.

@@

A stationary, smooth kernel may generally be expressed as a Taylor series for the $\tau \ne 0$ component,
plus the white noise component:
\begin{equation}
k(\tau) = a_0 \delta(\tau) +  \sum_{j=0}^\infty c_j \vert\tau\vert^j.
\end{equation}
Note that this encompasses all smooth, time-stationary kernels, and generally $c_1 \le 0$ as
correlated noise initially declines with separation in time.  We wish to approximate this
as the sum of damped cosine kernels, and since this general kernel is symmetric in $\tau$, we
can carry out the approximation in terms of sums of damped cosine kernel components
\begin{equation}
K_j(\tau,a_j,\beta_{R,j},\beta_{I,j}) = a_j \exp{(\beta_{R,j} \tau)}\cos{(\beta_{I,j} \tau)},
\end{equation}
which has a Taylor series expansion:
\begin{eqnarray}
\exp{(\beta_R \tau)}\cos{(\beta_I \tau)} &=& 1 - \beta_R \tau + \frac{1}{2}\left(\beta_R^2 - \beta_I^2\right)\tau^2 \\c &-&
\frac{1}{6} \left(\beta_R^3-3\beta_R\beta_I^2\right)\tau^3 + \frac{1}{24} \left(\beta_R^4+\beta_I^4-6\beta_R^2\beta_I^2\right) \tau^4+ O(\tau^5),
\end{eqnarray}
which may be continued to any order in $\tau$ as necessary.
If an approximation to $k(\tau)$ is desired in a Taylor series to some high order $N$, this
requires combining $N/2$ components of $K_j(\tau,a_j,\beta_{R,j},\beta_{I,j})$ with $\beta_{R,j}$ and $\beta_{I,j}$
chosen to match the coefficients $c_j$.  Note that for $N=1$ we only require a damped exponential
kernel with $\beta_I = 0$.

@@

\item {\bf Exponential kernel.} This kernel, $k(\tau) = e^{-\vert \tau/\tau_0\vert}$, is exactly

@@

\item {\bf Cosine kernel.} The cosine kernel, $k(t) = \cos{\left(\frac{2\pi}{P} \vert \tau \vert\right)}$,

@@

Precition/interpolation involves constructing a matrix of covariances
between the (noisy) training set and the times at which we wish to
predict or interpolate data points.   If our training times are given by
${\bf t}$, while the predicted times are given by ${\bf t}_*$, then
mean values at the predicted times are given by (RW 2.23):
\begin{eqnarray}
\bar f_{*,k} &=& \sum_j k(t_{*,k},t_j) b_j,\\
&=& \sum_j \sum_p \alpha_p e^{-\beta_p \vert t_{*,k}-t_j\vert} b_j,\\
&=& \sum_p \alpha_p \sum_j e^{-\beta_p \vert t_{*,k}-t_j\vert} b_j,\\
y_k &=& \sum_j k(t_j,t_k) b_j,
\end{eqnarray}
where the latter equation may be solved for ${\bf b}$ using the extended
matrix formalism.  Let the length of ${\bf t}_*$ be $M$; then, the
matrix $k({\bf t}_*,{\bf t})$ has a size $M \times N$, and so a total
of $MN$ multiplications is required to obtain the predicted values.
If $M$ is large, this can become quickly prohibitive.  It turns out
that the structure of the sum of exponential kernels may be exploited
to obtain the predicted mean in of order $p(M+N)+p^2N$ multiplications,
i.e.\ a linear scaling with the total number of data points, which can
be much fewer if $M$ is large.

@@

The variance may be handled in a similar manner, but with $p^2$ coefficients
to recursively update. For example, if $t_{*,k} > t_j \forall j,k$ then:
\begin{eqnarray}
cov(f_{*,k}) &=&  k(t_{*,k},t_{*,k}) - \sum_p \sum_q \alpha_p \alpha_q e^{-\beta_p(t_{*,k}-{\bf t}^T)} k({\bf t},{\bf t})^{-1} e^{-\beta_q (t_{*,k}-{\bf t})},\\
&=& k(0) - \sum_p \sum_q \alpha_p \alpha_q \zeta_{p,q,k},\\
\zeta_{p,q,k} &=&   e^{-\beta_p(t_{*,k}-{\bf t}^T)} k({\bf t},{\bf t})^{-1} e^{-\beta_q (t_{*,k}-{\bf t})},\\
&=&  e^{-(\beta_p+\beta_q)(t_{*,k}-t_{*,k-1})} \zeta_{p,q,k-1}.
\end{eqnarray}
The recursive updates to the $\zeta_{p,q,k}$ parameters can be made at each step.
\end{itemize}

@@

Here we adapt the generalized Rybicki-Press (GRP) algorithm to handle periodic kernels.  We
follow closely the notation and equations given in \citet{Ambikasaran:2015}.  We start
with the observation that a Lorentzian correlation function may be decomposed into the sum of two
exponential functions:
\begin{eqnarray}
k(\tau) &=& \alpha e^{-\beta_R \tau} \cos{(\omega \tau)}\\
 &=& \frac{1}{2}\alpha \left[e^{-\beta \tau}+e^{-\beta^* \tau}\right],
\end{eqnarray}
where $\beta = \beta_R + i\beta_I$ is the
complex kernel parameter (i.e.\ $\beta_R = Re(\beta)$, $\beta_I = Im(\beta)$), $\omega = -\beta_I = 2\pi/P$
is the frequency of oscillation of the kernel with period $P$, and $x^*$ is the complex conjugate of $x$.

@@

\begin{figure}[!htbp]
\begin{center}
\includegraphics[scale=0.175]{./twoterm.pdf}
\hspace{0.2in}
\includegraphics[scale=1.0]{./colorcode.pdf}
\end{center}
\caption{Pictorial description of the extended sparse matrix where $N=10$, $p_0=0$, and $p=2$, following
the example shown in \citet{Ambikasaran:2015}.}
\figurelabel{matrix_structure}
\end{figure}
